# `fertile`: My Contribution To Addressing Reproducibility {#my-solution}

## Understanding The Gaps In Existing Reproducibility Solutions

Although the current state of reproducibility in academia is quite poor, it is not an impossible challenge to overcome. The relative simplicity of addressing reproducibility, particularly when compared with replicability, makes it an ideal candidate for solution-building. Although significant progress on addressing reproducibility on a widespread scale is a long-term challenge, impactful forward progress--if on a smaller scale--can be achieved in the short-term. 

As we have seen, software developers, data scientists, and educators around the world have realized this potential, taking steps to help address the current crisis of reproducibility. Journals have put in place guidelines for authors, statisticians have developed `R` packages that help structure projects in a reproducible format, and educators have begun integrate reproducibility exercises into their courses.

However, many of these attempts to address reproducibility have significant drawbacks associated with them. We have already explored the issues with journal policies, both for authors and reviewers, in-depth. In this section, we will consider the education and software solutions and their associated challenges.

### In Education

The two primary concerns about the integration of reproducibility in data science curricula revolve around time and difficulty.

As noted previously, the primary mode of teaching reproducibility is through the assignment of replication studies where students must take an existing study and go through the process of reproducing it themselves, including contacting the author for all necessary materials, rerunning code and analysis, and problem-solving when issues almost certainly come up. 

In addition to the time required for the professor to collect all of the studies that students will be working on, the inclusion of such an assignment places a significant burden on educators by taking up time where they could be teaching other important material. Replication studies, if done correctly, can take weeks for students to successfully complete. The choice to give such assignments is therefore associated with a significant opportunity cost which many professors are unwilling to take.

Additionally, both replication studies assigned in class and replication workshops outside of normal coursework require a working knowledge of how to successfully complete and understand research. This makes them inaccessible to individuals who are still in their undergraduate career and may not yet have had an opportunity to conduct research or those who are studying in non-research-focused technical programs.

In order to reach the widest variety of students possible, it is necessary to develop a new method of teaching reproducibility that is neither time consuming nor dependent on a prior understanding of the research process.

### In Software

Previously, we considered several different types of software solutions: packages designed for users of `R` and continuous integration programs that can be used alongside a variety of coding languages. Although these solutions have their advantages, they also have significant drawbacks in terms of their ability to address reproducibility on a widespread scale.

Many of the packages designed for `R` are incredibly narrow in scope, with each effectively addressing a small component of reproducibility: file structure, modularization of code, version control, etc. They often succeed in their area of focus, but at the cost of accessibility to a wider audience. Their functions are often quite complex to use, and many steps must be completed to achieve the required reproducibility goal. This cumbersome nature means that most reproducibility packages currently available are not easily accessible to users with minimal `R` experience, nor particularly useful to those looking for quick and easy reproducibility checks. The significant learning curve associated with them can also detract potential users who may be interested in reproducibility but not willing to dedicate an extensive amount of time to understanding the intricacies of software operation.

Due to their generalized design, Continuous Integration tools do not face the same issues with narrowness or complexity that `R` packages struggle with. However, this generalizability provides its own additional challenge. Since Continuous Integration tools are designed to be accessible to a wide variety of users with different coding preferences, they are not particularly customizable and lack the ability to address features specific to certain programming languages. 

Neither of these different software solutions appear to adequately address the challenge of reproducibility. In order to be the most effective, a piece of software must instead:

1) Be simple, with a small library of functions/tools that are straightforward to use.
2) Be accessible to a variety of users, with a relatively small learning curvey.
3) Be able to address a wide variety of aspects of reproducibility, rather than just one or two key issues.
4) Have features specific to a particular coding language that can address that language's unique challenges.
5) Be customizable, allowing users to choose for themselves which aspects of reproducibility they want to focus on.

##  `fertile`, An R Package Creating Optimal Conditions For Reproducibility

What if it were possible to address the existing issues with both educational and software reproducibility solutions simultaneously? 

That is where my work comes in. In an attempt to produce meaningful change in the field of reproducibility, I have been developing `fertile`, a software package designed for `R` which helps users create optimal conditions for achieving reproducibility in their projects. 

`fertile` attempts to address the gaps in existing reproducbility solutions by combining software and education in one product. The package provides a set of simple, easy-to-learn tools that, rather than focus intensely on a specific area like other software programs, provide some information about a wide variety of aspects influencing reproducibility. It is also designed to be incredibly flexible, offering benefits to users at any stage in the data analysis workflow and providing users with the option to select which aspects of reproducibility they want to focus on. 

`fertile` also contains several`R`-specific features, which address certain aspects of reproducibility that can be missed by external project development tools. It is designed primarily to be used on data analyses organized as `R` Projects (i.e. directories containing an `.Rproj` file) and contains several associated features to ensure that the project structure meets the standards discussed in the `R` community.

In addition, `fertile` is designed to be educational, teaching its users about the components of reproducibility and how to achieve them in their work. The package provides users with detailed reports on the aspects of reproducibility where their projects fell short, identifying the root causes and, in many cases, providing a recommended solution.

`fertile` is structured in such a way as to be understandable and operable to individuals of any skill level, from students in their first undergraduate data science course to experienced PhD statisticians. The majority of its tools can be accessed in only a handful of functions with minimal required arguments. This simplicity makes the process of achieving and learning about reproducibility accessible to a wide audience in a way that complex software programs or graduate courses requiring an advanced knowledge of research methods do not.

Reproducibility is significantly easier to achieve when all of the tools necessary to do so are located in one place. `fertile` provides this optimal all-inclusive structure, addressing all six of the primary components of reproducibility discussed in Chapter 1. We will consider `fertile`'s treatment of each of these components in turn:

### Component 1: Accessibility of Project Files

  - Data (raw and/or processed)
  - Metadata
  - Code
  - Related Software
  
What we *have*:

* proj_report() gives the files present in the directory

What we *need*:

* can we report a checklist of the types of files we have available?
* it should check for the existence of a code file and data file and also readme

### Component 2: Organized Project Structure

  - Separate folders for different file types.
  - No extraneous files.
  - Minimal clutter.
  
`fertile` provides a wide variety of features for managing the file system of a project. Nine of the package's fifteen primary reproducibility checks relate to file structure.

Two of these are focused on the `R project` aspect of the file system. `has_proj_root()` ensures that there is a single `.Rproj` file indicating a clear root directory for the project, while `has_no_nested_proj_root()` ensures that there are no sub-projects within. The recognition of a clear root directory is necessary to allow for file structure analysis and project restructuring as it provides a baseline directory to define relative file paths from.

Six of the major checks, whose names begin with `has_tidy_` focus on file clutter. They check to make sure that no audio/video, image, source, raw data, .rda, or .R files are found in the root directory of the project.
  
  
What we *have*:

* has_tidy_media/raw_data/images/code/data/scripts
* has_proj_root/has_no_nested_proj_root
* has_only_used_files
* proj_move_files (and associated suggestions in proj_analyze())

What we *need*:

* nothing!


### Component 3: Documentation

  - Files are clearly named, preferably in a way where the order in which they should be run is clear.
  - A README is present.
  - Code contains comments.
  - Software dependencies are noted.
  
What we *have*:

* has_readme
* has_clear_build_chain
* list of packages in proj_analyze/report
* proj_pkg_script install script generator

What we *need*:

  - possibly a makefile generator? or improvements to order checking?
  - a way to note the sessioninfo and package numbers of a project
  - check if code contains comments

### Component 4: File Paths

  - No absolute paths.
  - No paths leading to locations outside of a project's directory.
  - Only relative paths, pointing to locations within a project's directory, are permitted.

What we *have*:

* proj_analyze_paths
* check_path

What we *need*:

* Nothing!

### Component 5: Randomness

  - If randomness is used in code, a seed must also be set.
  
What we *have*:

* has_no_randomness

What we *need*:

* Nothing!

### Component 6: Readability and Style

  - Though not mentioned in the sources described previously, it is also important that code be written in a coherent style. This is because code that conforms to a style guide or is written in a consistent dialect is easier to read, simplifying the process of following a researcher's work from beginning to end (@hermans2017programming). 

What we *have*:

* has_no_lint

What we *need*:

* nothing?? Ben's thoughts might be good here
  





Much of the available literature focuses on file structure, organization, and naming, and `fertile`'s features are consistent with this. @marwick2018packaging provide the framework for file structure that `fertile` is based on: a structure similar to that of an `R` package (@coreteam-extensions, @hadley-packages), with an `R` folder, as well as `data`, `data-raw`, `inst`, and `vignettes`.


 Once an `R` Project is created, `fertile` provides benefits throughout the data analysis process, both during development as well as after the fact. `fertile` achieves this by operating in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems).


### Proactive Use

Proactively, the package identifies potential mistakes as they are made by the user and outputs an informative message as well as a recommended solution. For example, `fertile` catches when a user passes a potentially problematic file path---such as an absolute path, or a path that points to a location outside of the project directory---to a variety of common input/output functions operating on many different file types.

\footnotesize

```{r, echo = FALSE, message = FALSE}
library(stringr)
wrapoutput <- function(output, width = 80) {
 message(paste(strwrap(output, width = width), collapse = "\n"))
}
```

```{r write_csv, include=FALSE}
readr::write_csv(mtcars, "~/Desktop/my_data.csv")
```

```{r fertile, message = FALSE, error = TRUE}
library(fertile)
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
read.csv("../../../Desktop/my_data.csv")
```

\normalsize

`fertile` is even more aggressive with functions (like `setwd()`) that are almost certain to break reproducibility, causing them to throw errors that prevent their execution and providing recommendations for better alternatives.

\footnotesize

```{r setwd, message=FALSE, error = TRUE}
setwd("~/Desktop")
```

\normalsize

These proactive warning features are activated immediately after attaching the `fertile` package and require no additional effort by the user.

### Retroactive Use

Retroactively, `fertile` analyzes potential obstacles to reproducibility in an RStudio Project (i.e., a directory that contains an `.Rproj` file). The package considers several different aspects of the project which may influence reproducibility, including the directory structure, file paths, and whether randomness is used thoughtfully.

The end products of these analyses are reproducibility reports summarizing a project's adherence to reproducibility standards and recommending remedies for where the project falls short. For example, `fertile` might identify the use of randomness in code and recommend setting a seed if one is not present.

Users can access the majority of `fertile`'s retroactive features through two primary functions, `proj_check()` and `proj_analyze()`. 

The `proj_check()` function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that do not have a call to `set.seed()` present. A full list is provided below:

\footnotesize

```{r list-checks, tidy.opts=list(width.cutoff=60)}
list_checks()
```

\normalsize

Subsets of the fifteen tests can be invoked using the `tidyselect` helper functions (@R-tidyselect) in combination with the more limited `proj_check_some()` function.

\footnotesize

```{r set-project}
proj_dir <- "project_miceps"
```

```{r echo=FALSE}
bad_file <- fs::path(proj_dir, "install_proj_packages.R")
if (file.exists(bad_file)) {
  fs::file_delete(bad_file)
}

```

```{r proj_check_some, warning = FALSE, tidy.opts=list(width.cutoff=60)}
proj_check_some(proj_dir, contains("paths"))
```

\normalsize

Each test can also be run individually by calling the function matching its check name.

The `proj_analyze()` function creates a report documenting the structure of a data analysis project. This report contains information about all packages referenced in code, the files present in the directory and their types, suggestions for moving files to create a more organized structure, and a list of reproducibility-breaking file paths used in code.

\footnotesize

```{r proj_analyze, tidy.opts=list(width.cutoff=60)}
proj_analyze(proj_dir)
```

\normalsize

### Logging

`fertile` also contains logging functionality, which records commands run in the console that have the potential to affect reproducibility, enabling users to look at their past history at any time. The package focuses mostly on package loading and file opening, noting which function was used, the path or package it referenced, and the timestamp at which that event happened. Users can access the log recording their commands at any time via the `log_report()` function:

\footnotesize

```{r read-csv, message=FALSE, include=FALSE}
library(purrr)
library(forcats)
read_csv(fs::path(proj_dir, "mice.csv"))
```

```{r log-report}
log_report()
```

\normalsize

The log, if not managed, can grow very long over time. For users who do not desire such functionality, `log_clear()` provides a way to erase the log and start over.

```{r log-clear, include=FALSE}
log_clear()
log_report()
```

### Utility Functions

`fertile` also provides several useful utility functions that may assist with the process of data analysis. 

### File Path Management

The `check_path()` function analyzes a vector of paths (or a single path) to determine whether there are any absolute paths or paths that lead outside the project directory.

\footnotesize

```{r path checks, error = TRUE, message = FALSE}
# Path inside the directory
check_path("project_miceps")

# Absolute path (current working directory)
check_path(getwd())

# Path outside the directory
check_path("../fertile.Rmd")
```

\normalsize


### File Types

There are several functions that can be used to check the type of a file: 

\footnotesize

```{r is_file}
is_data_file(fs::path(proj_dir, "mice.csv"))
is_image_file(fs::path(proj_dir, "proteins_v_time.png"))
is_text_file(fs::path(proj_dir, "README.md"))
is_r_file(fs::path(proj_dir, "analysis.Rmd"))
```

\normalsize

### Temporary Directories

The `sandbox()` function allows the user to make a copy of their project in a temporary directory. This can be useful for ensuring that projects run properly when access to the local file system is removed.

\footnotesize

```{r sandbox-1, results = 'hide'}
proj_dir
fs::dir_ls(proj_dir) %>% head(3)
```


```{r sandbox-2, echo = FALSE}
text <- fs::dir_ls(proj_dir) %>% head(3)
wrapoutput(str_wrap(text), 57)
```

```{r, sandbox-3, results = 'hide'}
temp_dir <- sandbox(proj_dir)
temp_dir
fs::dir_ls(temp_dir) %>% head(3)
```

```{r sandbox-4, echo = FALSE}
text <- fs::dir_ls(temp_dir) %>% head(3)
wrapoutput(str_wrap(text), 57)
```

\normalsize

### Managing Project Dependencies

One of the challenges with ensuring that work is reproducible is the issue of dependencies. Many data analysis projects reference a variety of `R` packages in their code. When such projects are shared with other users who may not have the required packages downloaded, it can cause errors that prevent the project from running properly. 

The `proj_pkg_script()`} function assists with this issue by making it simple and fast to download dependencies. When run on an `R` project directory, the function creates a `.R` script file that contains the code needed to install all of the packages referenced in the project, differentiating between packages located on CRAN and those located on GitHub.

\footnotesize

```{r proj_pkg_script, eval=TRUE, cache=TRUE, linewidth = 60}
install_script <- proj_pkg_script(proj_dir)
cat(readChar(install_script, 1e5))
```



\normalsize

## How `fertile` Works


Much of the functionality in `fertile` is achieved by writing `shims` **link to wikipedia page here**. `fertile`'s shimmed functions intercept the user's commands and perform various logging and checking tasks before executing the desired function. Our process is:

1. Identify an `R` function that is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., `read_csv()` from `readr`) are ideal candidates.

2. Create a function in `fertile` with the same name that takes the same arguments (and always the dots `...`).

3. Write this new function so that it: 
  a) captures any arguments, 
  b) logs the name of the function called, 
  c) performs any checks on these arguments, and 
  d) calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
  
Most shims are quite simple and look something like what is shown below for `read_csv()`. 

\footnotesize

```{r}
fertile::read_csv
```

\normalsize

`fertile` shims many common functions, including those that read in a variety of data types, write data, and load packages. This works both proactively and retroactively, as the shimmed functions written in `fertile` are activated both when the user is coding interactively and when a file containing code is rendered.

In order to ensure that the `fertile` versions of functions ("shims") always supersede ("mask") their original namesakes when called, `fertile` uses its own shims of the `library` and `require` functions to  manipulate the `R` `search` path so that it is always located in the first position. In the `fertile` version of `library()`, we detach `fertile` from the search path, load the requested package, and then re-attach `fertile`. This ensures that when a user executes a command, `R` will check `fertile` for a matching function before considering other packages. While it is possible that this shifty behavior could lead to unintended consequences, our goal is to catch a good deal of problems before they become problematic. Users can easily disable `fertile` by detaching it, or not loading it in the first place. 


## `fertile` in Practice: Experimental Results From Smith College Student Use

`fertile` is designed to: 1) be simple enough that users with minimal `R` experience can use the package without issue, 2) increase the reproducibility of work produced by its users, and 3) educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.

To test `fertile`'s effectiveness, we began an initial randomized control trial of the package on an introductory undergraduate data science course at Smith College in Spring 2020 **ADD FOOTNOTE** (This study was approved by Smith College IRB, Protocol #19-032).

The experiment was structured as follows:


1.Students are given a form at the start of the semester asking whether they consent to participate in a study on data science education. In order to successfully consent, they must provide their system username, collected through the command `Sys.getenv("LOGNAME")`. To maintain privacy the results are then transformed into a hexadecimal string via the `md5()` hashing function. 

2. These hexadecimal strings are then randomly assigned into equally sized groups, one experimental group that receives the features of `fertile` and one group that receives a control.

3. The students are then asked to download a package called `sds192` (the course number and prefix), which was created for the purpose of this trial. It leverages an `.onAttach()` function to scan the `R` environment and collect the username of the user who is loading the package and run it through the same hashing algorithm as used previously. It then identifies whether that user belongs to the experimental or the control group. Depending on the group they are in, they receive a different version of the package.

4. The experimental group receives the basic `sds192` package, which consists of some data sets and `R` Markdown templates necessary for completing homework assignments and projects in the class, but also has `fertile` installed and loaded silently in the background. The package's proactive features are enabled, and therefore users will receive warning messages when they use absolute or non-portable paths or attempt to change their working directory. The control group receives only the basic `sds192` package, including its data sets and `R` Markdown templates. All students from both groups then use their version of the package throughout the semester on a variety of projects.

5. Both groups are given a short quiz on different components of reproducibility that are intended to be taught by `fertile` at both the beginning and end of the semester. Their scores are then compared to see whether one group learned more than the other group or whether their scores were essentially equivalent. Additionally, for every homework assignment submitted, the professor takes note of whether or not the project compiles successfully.


Based on the results, we hope to determine whether `fertile` was successful at achieving its intended goals. A lack of notable difference between the *experimental* and *control* groups in terms of the number of code-related questions asked throughout the semester would indicate that `fertile` achieved its goal of simplicity. A higher average for the *experimental* group in terms of the number of homework assignments that compiled successfully would indicate that `fertile` was successful in increasing reproducibility. A greater increase over the semester in the reproducibility quiz scores for students in the *experimental* group compared with the *control* group would indicate that `fertile` achieved its goal of educating users on reproducibility. Success according to these metrics would provide evidence showing `fertile`'s benefit as tool to help educators introduce reproducibility concepts in the classroom.

