# `fertile`: My Contribution To Addressing Reproducibility {#my-solution}

## Understanding The Gaps In Existing Reproducibility Solutions

Although the current state of reproducibility in academia is quite poor, it is not an impossible challenge to overcome. The relative simplicity of addressing reproducibility, particularly when compared with replicability, makes it an ideal candidate for solution-building. Although significant progress on addressing reproducibility on a widespread scale is a long-term challenge, impactful forward progress--if on a smaller scale--can be achieved in the short-term. 
As we have seen, software developers, data scientists, and educators around the world have realized this potential, taking steps to help address the current crisis of reproducibility. Journals have put in place guidelines for authors, statisticians have developed `R` packages that help structure projects in a reproducible format, and educators have begun integrate reproducibility exercises into their courses.

However, many of these attempts to address reproducibility have significant drawbacks associated with them. We have already explored the issues with journal policies, both for authors and reviewers, in-depth. In this section, we will consider the education and software solutions and their associated challenges.

### In Education

The two primary concerns about the integration of reproducibility in data science curricula revolve around time and difficulty.

As noted previously, the primary mode of teaching reproducibility is through the assignment of replication studies where students must take an existing study and go through the process of reproducing it themselves, including contacting the author for all necessary materials, rerunning code and analysis, and problem-solving when issues almost certainly come up. 

In addition to the time required for the professor to collect all of the studies that students will be working on, the inclusion of such an assignment places a significant burden on educators by taking up time where they could be teaching other important material. Replication studies, if done correctly, can take weeks for students to successfully complete. The choice to give such assignments is therefore associated with a significant opportunity cost which many professors are unwilling to take.

Additionally, both replication studies assigned in class and replication workshops outside of normal coursework require a working knowledge of how to successfully complete and understand research. This makes them inaccessible to individuals who are still in their undergraduate career and may not yet have had an opportunity to conduct research or those who are studying in non-research-focused technical programs.

In order to reach the widest variety of students possible, it is necessary to develop a new method of teaching reproducibility that is neither time consuming nor dependent on a prior understanding of the research process.

### In Software

Previously, we considered several different types of software solutions: packages designed for users of `R` and continuous integration programs that can be used alongside a variety of coding languages. Although these solutions have their advantages, they also have significant drawbacks in terms of their ability to address reproducibility on a widespread scale.

Many of the packages designed for `R` are incredibly narrow in scope, with each effectively addressing a small component of reproducibility: file structure, modularization of code, version control, etc. They often succeed in their area of focus, but at the cost of accessibility to a wider audience. Their functions are often quite complex to use, and many steps must be completed to achieve the required reproducibility goal. This cumbersome nature means that most reproducibility packages currently available are not easily accessible to users with minimal `R` experience, nor particularly useful to those looking for quick and easy reproducibility checks. The significant learning curve associated with them can also detract potential users who may be interested in reproducibility but not willing to dedicate an extensive amount of time to understanding the intricacies of software operation.

Due to their generalized design, Continuous Integration tools do not face the same issues with narrowness or complexity that `R` packages struggle with. However, this generalizability provides its own additional challenge. Since Continuous Integration tools are designed to be accessible to a wide variety of users with different coding preferences, they are not particularly customizable and lack the ability to address features specific to certain programming languages. 

Neither of these different software solutions appear to adequately address the challenge of reproducibility. In order to be the most effective, a piece of software must instead:

1) Be simple, with a small library of functions/tools that are straightforward to use.
2) Be accessible to a variety of users, with a relatively small learning curvey.
3) Be able to address a wide variety of aspects of reproducibility, rather than just one or two key issues.
4) Have features specific to a particular coding language that can address that language's unique challenges.
5) Be customizable, allowing users to choose for themselves which aspects of reproducibility they want to focus on.

##  `fertile`, An R Package Creating Optimal Conditions For Reproducibility

What if it were possible to address the existing issues with both educational and software reproducibility solutions simultaneously? 

That is where my work comes in. In an attempt to produce meaningful change in the field of reproducibility, I have been developing `fertile`, a software package designed for `R` which helps users create optimal conditions for achieving reproducibility in their projects. 

`fertile` attempts to address the gaps in existing reproducbility solutions by combining software and education in one product. The package provides a set of simple, easy-to-learn tools that, rather than focus intensely on a specific area like other software programs, provide some information about a wide variety of aspects influencing reproducibility. It is also designed to be incredibly flexible, offering benefits to users at any stage in the data analysis workflow and providing users with the option to select which aspects of reproducibility they want to focus on. 

`fertile` also contains several`R`-specific features, which address certain aspects of reproducibility that can be missed by external project development tools. It is designed primarily to be used on data analyses organized as `R` Projects (i.e. directories containing an `.Rproj` file) and contains several associated features to ensure that the project structure meets the standards discussed in the `R` community.

In addition, `fertile` is designed to be educational, teaching its users about the components of reproducibility and how to achieve them in their work. The package provides users with detailed reports on the aspects of reproducibility where their projects fell short, identifying the root causes and, in many cases, providing a recommended solution.

`fertile` is structured in such a way as to be understandable and operable to individuals of any skill level, from students in their first undergraduate data science course to experienced PhD statisticians. The majority of its tools can be accessed in only a handful of functions with minimal required arguments. This simplicity makes the process of achieving and learning about reproducibility accessible to a wide audience in a way that complex software programs or graduate courses requiring an advanced knowledge of research methods do not.

Reproducibility is significantly easier to achieve when all of the tools necessary to do so are located in one place. `fertile` provides this optimal all-inclusive structure, addressing all six of the primary components of reproducibility discussed in Chapter 1. We will consider `fertile`'s treatment of each of these components in turn:

### Component 1: Accessibility of Project Files

  - Data (raw and/or processed)
  - Metadata
  - Code
  - Related Software
  
### Temporary Directories

The `sandbox()` function allows the user to make a copy of their project in a temporary directory. This can be useful for ensuring that projects run properly when access to the local file system is removed.

\footnotesize
```{r, echo=FALSE}
library(tidyverse)
library(fertile)
```


```{r sandbox-1, results = 'hide'}
fs::dir_ls('project_miceps') %>% head(3)
```


```{r sandbox-2, echo = FALSE}
text <- fs::dir_ls('project_miceps') %>% head(3)
#wrapoutput(str_wrap(text), 57)
```

```{r, sandbox-3, results = 'hide'}
temp_dir <- sandbox('project_miceps')
temp_dir
fs::dir_ls(temp_dir) %>% head(3)
```

```{r sandbox-4, echo = FALSE}
text <- fs::dir_ls(temp_dir) %>% head(3)
#wrapoutput(str_wrap(text), 57)
```

\normalsize

  
  
  
What we *have*:

* proj_report() gives the files present in the directory
* Every time a project is rendered, a session info file is created
* proj_pkg_script() for dependencies
* sandbox

What we *need*:

* can we report a checklist of the types of files we have available?
* it should check for the existence of a code file and data file and also readme

### Component 2: Organized Project Structure

`fertile` provides a wide variety of features for managing the file system of a project. Nine of the package's fifteen primary reproducibility checks relate to file structure.

Two of these are focused on the `R project` aspect of the file system. `has_proj_root()` ensures that there is a single `.Rproj` file indicating a clear root directory for the project, while `has_no_nested_proj_root()` ensures that there are no sub-projects within. The recognition of a clear root directory is necessary to allow for file structure analysis and project restructuring as it provides a baseline directory to define relative file paths from.

Six of the major checks, whose names begin with `has_tidy_` focus on file clutter, addressing one of the components of a clean file structure. They check to make sure that no audio/video, image, source, raw data, .rda, or .R files are found in the root directory of the project.

The last check that focuses on the file system is `has_only_used_files()`. One of the more complicated checks in `fertile`, this function checks to make sure that there are no extraneous files serving no purpose (that are not "used") included alongside the analysis. For this function to work properly, a clear definition of which files are considered as "used" is needed. In `fertile`, that definition includes the following:

* `.R`,`.Rmd`, and `.Rproj` files
* README files
* Data/text/image files whose file path is referenced, either as an input or output, in any of the `.R` or `.Rmd` files
* Outputs created by knitting any `.Rmd` files (`.html`, `.pdf`, or `.docx` files with the same name as the `.Rmd`)
* Files created by `fertile`. In an effort to capture information about dependencies, `fertile` creates a text file capturing the session information when a project is run and provides an option to generate an `.R` script that can install all of the packages referenced in a project's code. These files are considered "used."


```{r, message = FALSE}
library(fertile)
```

```{r}
# List the files in the directory
fs::dir_ls("project_miceps")

# Check to see which are "used"
has_only_used_files("project_miceps")

```

There are also several functions that summarize and reshape the entire project structure. `proj_analyze_files()` lists all of the files in the project, along with their size, type, and relative path within the directory. 

```{r}
proj_analyze_files("project_miceps")
```

`proj_suggest_moves()` provides suggestions for how the files should be reorganized, if at all. 

```{r}
files <- proj_analyze_files("project_miceps")

proj_suggest_moves(files)
```


These suggestions are based on achieving the optimal file structure design for reproducibility, argued by @marwick2018packaging to be that of an `R` package (@coreteam-extensions, @hadley-packages), with an `R` folder, as well as `data`, `data-raw`, `inst`, and `vignettes`. Such a structure keeps all of the files separated and in clearly labeled directories where they are easy to find. Additionally, the extension of the `R` package structure to data analysis projects promotes standardization of file structure within the R community. `proj_move_files()` executes these suggestions, building a package structure and sorting files into the correct folders by type.

Combined together, all of these functions make it simple to ensure that projects fall under a standard, simple, and reproducible structure with minimal clutter.


### Component 3: Documentation

  - Files are clearly named, preferably in a way where the order in which they should be run is clear.
  - A README is present.
  - Code contains comments.
  - Software dependencies are noted.
  
High quality documentation, including the presence of a README file, comments explaining the code, a list of software packages an analysis is dependent on, and a method in which to understand which order to run code files in, is incredibly important to reproducibility. Without written guidance, individuals looking to reproduce results may not understand how to take all of the project components and combine them in the right way to produce the desired outcome. Additionally, code used without the proper dependencies and software versions, even if it is perfectly functional and correctly written, will often result in errors when compiled. 

`fertile` contains a variety of functions to ensure that projects are well documented.

The `has_readme()` function ensures that a README file is included alongside a project.

```{r}
has_readme("project_miceps")
```


`has_clear_build_chain()` looks to see if the project includes a `makefile`/`drakefile` (make-style file generated by the `Drake` package in `R`) or, if multiple code files are included, whether they are numbered. Any of these conditions being true would indicate that the order in which to run different pieces of code is clear. `project_miceps`, since it only contains one code file, passes this check easily:

```{r}
has_clear_build_chain("project_miceps")
```

There are also several features focused on software dependencies. Every time the code contained within a project is rendered/run by `fertile`, the package generates a file capturing the `sessionInfo()` just after the code was run. This file contains information about the R version in which the code was run, the list of packages that were loaded, and their specific versions. 

The dependency information from `project_miceps` looks as follows:

```{r}
session_info <- "project_miceps/software-versions.txt"
cat(readChar(session_info, 1e5))
```

The `proj_pkg_script()` makes it simple and fast to download these dependencies. When run on an `R` project directory, the function creates a `.R` script file that contains the code needed to install all of the packages referenced in the project, differentiating between packages located on CRAN and those located on GitHub.

\footnotesize

```{r eval=TRUE, cache=TRUE, linewidth = 60}
install_script <- proj_pkg_script("project_miceps")
cat(readChar(install_script, 1e5))
```

\normalsize

What we *need*:

  - possibly a makefile generator? or improvements to order checking?
  

### Component 4: File Paths

`fertile` contains a variety of features designed to address issues with file paths. Several of these features happen proactively, warning users that they are entering a non-reproducible file path as it happens, while others can be accessed after a project has already been written.

Proactively, `fertile` catches when an absolute path, or one leading outside of the project directory, is referenced in an input or output function and outputs a warning message. This interactive behavior is one of the educational features of the package, as it immediately provides an informative message explaining what went wrong, giving the programmer an opportunity to learn and course-correct from there.


\footnotesize

```{r, echo = FALSE, message = FALSE}
library(stringr)
wrapoutput <- function(output, width = 80) {
 message(paste(strwrap(output, width = width), collapse = "\n"))
}
```

```{r write_csv, include=FALSE}
readr::write_csv(mtcars, "~/Desktop/my_data.csv")
```

```{r fertile, message = FALSE, error = TRUE}
library(fertile)
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
read.csv("../../../Desktop/my_data.csv")
```

\normalsize

`fertile` is particularly strict with the function `setwd()`, as its use is almost certain to break reproducibility. Calls to `setwd()` are blocked from executing, and users must conduct a manual override, via the `danger()` function, to change their working directory.

\footnotesize

```{r setwd, message=FALSE, error = TRUE}
getwd()
setwd("~/Desktop")
getwd()
```

`fertile` also contains functionality that analyzes code for path information after it is written. For example, `proj_analyze_paths` goes through all of the paths used in a project's `.R` and `.Rmd` files and produces a report detailing which ones failed to meet reproducibility standards, explaining the problem, and providing a solution.

```{r}
proj_analyze_paths('project_miceps')
```

Paths can also be checked individually, or in groups, using `check_path()`.

\footnotesize

```{r path checks, error = TRUE, message = FALSE}
# Path inside the directory
check_path("project_miceps")

# Absolute path (current working directory)
check_path(getwd())

# Path outside the directory
check_path("../README.md")
```

\normalsize

These functions, together, cover all of the reproducibility subcomponents involving file paths. Users are warned not to use absolute paths or paths pointing outside of the directory, both while they are coding and after the fact, and provided with a recommended solution to the problem if they do so. Additionally, users are prevented from using functions that will certainly break reproducibility and provided with safer alternatives. 

### Component 5: Randomness

The component of randomness is addressed by `fertile` in a reproducibility check function titled `has_no_randomness()`, which does the following: First, it checks code files to determine whether they have employed random number generation. This is confirmed by recording the random number generation seed prior to running the files, rendering the files, and then checking to see whether the seed has changed. If the seed has changed, that indicates that random number generation has occurred. Then, it determines whether whether any randomness present was semi-random (i.e. controlled and repeatable) or truly random, by checking whether the `set.seed()` function was called. If the randomness was semi-random, it is considered permissible and reproducible. Truly random code is flagged as dangerous.

Let us consider an example from `project_miceps`. The primary code file, `analysis.Rmd` contains the following code. In this code, we see that a data set is read in via `read_csv` and reformatted. Then, the `sample_frac()` function takes a random sample of 50% of the rows of the data.

```{r, message=FALSE, eval = FALSE}

mice <- read_csv("Blot_data_updated.csv") %>%
  transmute(time = `Time point`,
            sex = Sex,
            id = `Mouse #`,
            erb = `ERB Normalized Final`,
            era = `ERA Normalized Final`,
            gper = `GPER Normalized Final`) %>%
  mutate(day = case_when(
    time == 0 ~ -1,
    time == 1 ~ 0,
    time == 2 ~ 1,
    time == 3 ~ 3,
    time == 4 ~ 5,
    time == 5 ~ 7
  ),
  type = ifelse(time == 0, "control", "treatment")
  )
mice_tidy <- mice %>%
  select(-time) %>%
  gather(key = "protein", value = "amount", -day, -sex, -id, -type) %>%
  mutate(protein_label = factor(protein,
    labels = c("paste(ER, alpha)", "paste(ER, beta)", "GPER")))
```

```{r, message=FALSE, echo = FALSE}
library(dplyr)
library(tidyr)
mice <- read_csv("project_miceps/Blot_data_updated.csv") %>%
  transmute(time = `Time point`,
            sex = Sex,
            id = `Mouse #`,
            erb = `ERB Normalized Final`,
            era = `ERA Normalized Final`,
            gper = `GPER Normalized Final`) %>%
  mutate(day = case_when(
    time == 0 ~ -1,
    time == 1 ~ 0,
    time == 2 ~ 1,
    time == 3 ~ 3,
    time == 4 ~ 5,
    time == 5 ~ 7
  ),
  type = ifelse(time == 0, "control", "treatment")
  )
mice_tidy <- mice %>%
  select(-time) %>%
  gather(key = "protein", value = "amount", -day, -sex, -id, -type) %>%
  mutate(protein_label = factor(protein,
   labels = c("paste(ER, alpha)", "paste(ER, beta)", "GPER")))
```


```{r, eval = FALSE}
sample_frac(mice, 0.5)
```

However, there is no seed set in the file. As a result, `has_no_randomness()` returns as a failure for the project. Had a seed been set, however, the project would have passed.

```{r}
has_no_randomness('project_miceps')
```

`fertile`'s randomness-centered feature helps analysts know that their use of randomness is controlled, ensuring that functions involving random number generation will always produce the same output each time they are run.

### Component 6: Readability and Style

Though not an absolutely necessary component of reproducibility, code style can have a significant impact on how easy it is to follow the steps being taken in an analysis. The use of consistent and highly-readable style in code greatly simplifies and speeds up the process of understanding a data analysis project and repeating the steps included within. 

`fertile` addresses code style via the function `has_no_lint()`. This function checks code for compatibility with RStudio Chief Scientist Hadley Wikham's "tidy" code format, which promotes the following best practices:

* Line length should not exceed 80 characters
* There should not be trailing whitespace
* All infix operators (`=`, `+`, `-`, `<-`, etc) should have spaces on either side
* All commas should be followed by spaces
* There is no code left in the file that is commented out
* `<-` should be used instead of `=` to assign variables
* Opening curly braces should never go on their own line and should always be followed by a new line
* There should always be a space between right parenthesis and an open curly brace
* Closing curly braces should always be on their own line, unless followed by an `else` statement

Any issues with incompatibility that are found by `has_no_lint()` appear in a window by the RStudio console. This window lists the style inconsistencies found in each code document, showing both the description of the error and the line number where it occurred for each issue. This window is interactive; double-clicking on an error brings the user to exact location in the code where it occurred.

Let us consider an example. `project_miceps` contains an `RMarkdown` file titled `analysis.Rmd`, which contains some code involved with a data analysis. Some of the code is in tidy style but not all of it is. 

For example, line 71, part of a `ggplot` call, contains the following code:

```{r, eval = FALSE}
# Line 71 of `project_miceps/analysis.Rmd'

geom_hline(aes(yintercept = estimate + 1.96*std.error, color = sex), linetype = 3) + 
  
```

For this line, `has_no_lint()` finds the following inconsistencies with tidy style:

**Line 71** Lines should not be more than 80 characters.  

**Line 71** Put spaces around all infix operators.  

**Line 71** Trailing whitespace is superfluous.    

The first comes up because the true line length, including spaces, is 84 characters. The second comes up because the `*` between `1.96` and `std.error` is not surrounded by spaces. The third is flagged because there is an empty space after the `+` at the end of the line.

An even less tidy piece of code is that found on line 189, written as follows:

```{r}
# Line 189 of `project_miceps/analysis.Rmd'

if (length(mice) > 1){ holder_var = 1 }

```

For this line, `has_no_lint()` finds the following errors:

**Line 189** There should be a space between right parenthesis and an opening curly brace.  

**Line 189** Opening curly braces should never go on their own line and should always be followed by a new line.  

**Line 189** Use <-, not =, for assignment.  

**Line 189** Closing curly braces should always be on their own line, unless it's followed by an else.  

The first issue flagged is the fact that `){`, in the middle of the line, has no space in the middle. The second issue is flagged because the code to execute when the `if` statement is true (`holder_var = 1`) is on the same line as the `if` statement, rather than on its own line. The third error occurs because `holder_var` is defined using an `=`. And the fourth occurs because the closing curly brace is on the same line as other code rather than by itself.

These informative error messages provided by `fertile` help teach users to use consistent, tidy style in their code. They do so while making the learning process incredibly simple, allowing users to click and see exactly which symbol or character caused the error. In addition to providing a detailed explanation of the was in which the provided code is not tidy, `fertile` also suggests a simple and fast solution for resolving the inconsistencies all at once, recommending that users apply the `usethis::use_tidy_style()` function to format their code automatically.

### User Customizability

`fertile` does not force its users into a box, instead allowing for a great deal of user choice in terms of which aspects of reproducibility to focus on. Users can run reproducibility tests at three different scales: 

1) Comprehensively, where all checks are run within a single function or two.
2) In groups, where functions focused on similar aspects of reproducibility are run together.
3) Individually, where only one reproducbility check is run at a time.

Users who want comprehensive behavior can access it through two primary functions, `proj_check()` and `proj_analyze()`. 

The `proj_check()` function runs all fifteen reproducibility tests in the package, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests, many of which were described in detail previously, include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that do not have a call to `set.seed()` present. A full list is provided below:

\footnotesize

```{r list-checks, tidy.opts=list(width.cutoff=60)}
list_checks()
```

\normalsize


The `proj_analyze()` function creates a report documenting the structure of a data analysis project, combining four key functions from `fertile` into one complete unit:

* `proj_analyze_packages()`, which captures a list of all packages referenced in the source code and which files they were referenced in
* `proj_analyze_files()`, which captures a list of all of the files located in the directory along with their type and size
* `proj_suggest_moves()`, which provides suggestions for how to reorganize files to create a more reproducible file structure
* `proj_analyze_paths()`, which captures a list of the non-reproducible file paths referenced in source code

\footnotesize

```{r proj_analyze, tidy.opts=list(width.cutoff=60)}
proj_analyze('project_miceps')
```

\normalsize

Together, `proj_check()` and `proj_analyze()` cover a significant portion of all six major components of reproducibility.

Users wanting to focus on groups of checks can do so using the `proj_check_some()` function.
`proj_check_some()` leverages helper functions from `tidyselect` (@R-tidyselect) to allow users to call groups of similar checks together. `tidyselect` helpers tell `fertile` to call only the checks whose names meet certain conditions. The helper functions that can be passed to `proj_check_some()` are:

* `starts_with()`: Starts with a prefix.
* `ends_with()`: Ends with a suffix.
* `contains()`: Contains a literal string.
* `matches()`: Matches a regular expression.

For example, a variety of checks in `fertile` focus on making sure the project has a "tidy" structure--essentially that there are not files cluttered together all in one folder. Users interested in checking their tidyness can do so all at once using `proj_check_some()` as follows:

\footnotesize


```{r echo=FALSE}
bad_file <- fs::path('project_miceps', "install_proj_packages.R")
if (file.exists(bad_file)) {
  fs::file_delete(bad_file)
}

```

```{r proj_check_some, warning = FALSE, tidy.opts=list(width.cutoff=60)}
proj_check_some("project_miceps", contains("tidy"))
```

\normalsize

Users might also attempt to call the two checks involving project roots together:

```{r}
proj_check_some("project_miceps", ends_with("root"))
```

Or, perhaps, they might want to run all the functions beginning with "has_only":

```{r}
proj_check_some("project_miceps", starts_with("has_only"))
```

If users do not want to run functions in groups, and prefer to run them individually, that option is also provided to them. Every check that makes up `check()` and every subcomponent of `project_analyze()`can be run individually. 


### Educational Features

Simply noting and correcting issues with reproducibility is not enough to produce lasting change in the scientific community. Data analysts and software users must also be educated on why their choices were not reproducible so that they do not fall victim to those mistakes again in the future, but also so that they can share their knowledge and experience with others in the scientific community.

`fertile` prioritizes this idea of reproducibility education throughout many of its functionalities.


 which records commands run in the console that have the potential to affect reproducibility, enabling users to look at their past history at any time. The package focuses mostly on package loading and file opening, noting which function was used, the path or package it referenced, and the timestamp at which that event happened. Users can access the log recording their commands at any time via the `log_report()` function:

\footnotesize

```{r read-csv, message=FALSE, include=FALSE}
library(purrr)
library(forcats)
read_csv(fs::path('project_miceps', "mice.csv"))
```

```{r log-report}
log_report()
```

\normalsize

The log, if not managed, can grow very long over time. For users who do not desire such functionality, `log_clear()` provides a way to erase the log and start over.

```{r log-clear, include=FALSE}
log_clear()
log_report()
```


### File Types

There are several functions that can be used to check the type of a file: 

\footnotesize

```{r is_file}
is_data_file(fs::path('project_miceps', "mice.csv"))
is_image_file(fs::path('project_miceps', "proteins_v_time.png"))
is_text_file(fs::path('project_miceps', "README.md"))
is_r_file(fs::path('project_miceps', "analysis.Rmd"))
```

\normalsize



## How `fertile` Works


Much of the functionality in `fertile` is achieved by writing `shims` **link to wikipedia page here**. `fertile`'s shimmed functions intercept the user's commands and perform various logging and checking tasks before executing the desired function. Our process is:

1. Identify an `R` function that is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., `read_csv()` from `readr`) are ideal candidates.

2. Create a function in `fertile` with the same name that takes the same arguments (and always the dots `...`).

3. Write this new function so that it: 
  a) captures any arguments, 
  b) logs the name of the function called, 
  c) performs any checks on these arguments, and 
  d) calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
  
Most shims are quite simple and look something like what is shown below for `read_csv()`. 

\footnotesize

```{r}
fertile::read_csv
```

\normalsize

`fertile` shims many common functions, including those that read in a variety of data types, write data, and load packages. This works both proactively and retroactively, as the shimmed functions written in `fertile` are activated both when the user is coding interactively and when a file containing code is rendered.

In order to ensure that the `fertile` versions of functions ("shims") always supersede ("mask") their original namesakes when called, `fertile` uses its own shims of the `library` and `require` functions to  manipulate the `R` `search` path so that it is always located in the first position. In the `fertile` version of `library()`, we detach `fertile` from the search path, load the requested package, and then re-attach `fertile`. This ensures that when a user executes a command, `R` will check `fertile` for a matching function before considering other packages. While it is possible that this shifty behavior could lead to unintended consequences, our goal is to catch a good deal of problems before they become problematic. Users can easily disable `fertile` by detaching it, or not loading it in the first place. 


## `fertile` in Practice: Experimental Results From Smith College Student Use

`fertile` is designed to: 1) be simple enough that users with minimal `R` experience can use the package without issue, 2) increase the reproducibility of work produced by its users, and 3) educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.

To test `fertile`'s effectiveness, we began an initial randomized control trial of the package on an introductory undergraduate data science course at Smith College in Spring 2020 **ADD FOOTNOTE** (This study was approved by Smith College IRB, Protocol #19-032).

The experiment was structured as follows:


1.Students are given a form at the start of the semester asking whether they consent to participate in a study on data science education. In order to successfully consent, they must provide their system username, collected through the command `Sys.getenv("LOGNAME")`. To maintain privacy the results are then transformed into a hexadecimal string via the `md5()` hashing function. 

2. These hexadecimal strings are then randomly assigned into equally sized groups, one experimental group that receives the features of `fertile` and one group that receives a control.

3. The students are then asked to download a package called `sds192` (the course number and prefix), which was created for the purpose of this trial. It leverages an `.onAttach()` function to scan the `R` environment and collect the username of the user who is loading the package and run it through the same hashing algorithm as used previously. It then identifies whether that user belongs to the experimental or the control group. Depending on the group they are in, they receive a different version of the package.

4. The experimental group receives the basic `sds192` package, which consists of some data sets and `R` Markdown templates necessary for completing homework assignments and projects in the class, but also has `fertile` installed and loaded silently in the background. The package's proactive features are enabled, and therefore users will receive warning messages when they use absolute or non-portable paths or attempt to change their working directory. The control group receives only the basic `sds192` package, including its data sets and `R` Markdown templates. All students from both groups then use their version of the package throughout the semester on a variety of projects.

5. Both groups are given a short quiz on different components of reproducibility that are intended to be taught by `fertile` at both the beginning and end of the semester. Their scores are then compared to see whether one group learned more than the other group or whether their scores were essentially equivalent. Additionally, for every homework assignment submitted, the professor takes note of whether or not the project compiles successfully.


Based on the results, we hope to determine whether `fertile` was successful at achieving its intended goals. A lack of notable difference between the *experimental* and *control* groups in terms of the number of code-related questions asked throughout the semester would indicate that `fertile` achieved its goal of simplicity. A higher average for the *experimental* group in terms of the number of homework assignments that compiled successfully would indicate that `fertile` was successful in increasing reproducibility. A greater increase over the semester in the reproducibility quiz scores for students in the *experimental* group compared with the *control* group would indicate that `fertile` achieved its goal of educating users on reproducibility. Success according to these metrics would provide evidence showing `fertile`'s benefit as tool to help educators introduce reproducibility concepts in the classroom.

