---
title: 'Addressing The Scientific Reproducibility Crisis Through Educational Software Integration'
author: 'Audrey M. Bertin'
date: 'May 2021'
institution: 'Smith College'
division: 'Statistical and Data Sciences'
advisor: 'Benjamin S. Baumer'
altadvisor: 'Albert Y. Kim'
department: 'Statistical and Data Sciences'
degree: 'Bachelor of Arts'
knit: bookdown::render_book
site: bookdown::bookdown_site

# This will automatically install the {remotes} package and {thesisdown}
# Change this to FALSE if you'd like to install them manually on your own.
params:
  'Install needed packages for {thesisdown}': True
  
# Remove the hashtag to specify which version of output you would like.
# Can only choose one at a time.
output:
  thesisdown::thesis_pdf: default 
  
#  thesisdown::thesis_gitbook: default         
#  thesisdown::thesis_word: default
#  thesisdown::thesis_epub: default

# If you are creating a PDF you'll need to write your preliminary content 
# (e.g., abstract, acknowledgements) below or use code similar to line 25-26 
# for the .RMD files. If you are NOT producing a PDF, delete or silence
# lines 25-39 in this YAML header.
abstract: '`r if(knitr:::is_latex_output()) paste(readLines(here::here("prelims", "00-abstract.Rmd")), collapse = "\n  ")`'
# If you'd rather include the preliminary content in files instead of inline
# like below, use a command like that for the abstract above.  Note that a tab 
# is needed on the line after the `|`.
acknowledgements: |
  This project would not have been possible without the guidance of Professor Ben Baumer, who helped inspire my interest in data science and mentored me throughout this project for the last two years. 
  I would also like to extend my thanks to Jenny Bryan and Hadley Wickham, whose advice has helped guide the development of several features in `fertile`, along with the authors of all of the R packages utilized by `fertile` to achieve its functionality. 
  Additionally, a huge thanks to the many introductory data science students who volunteered their time as part of a study to test the effectiveness of implementing `fertile` in the classroom. The help of all these Smithies was invaluable in ensuring that the overall `fertile` product was as user friendly and effective as possible.  
  
  
#dedication: |
#  You can have a dedication here if you wish. 
#preface: |
#  This is an example of a thesis setup to use the reed thesis document class 
#  (for LaTeX) and the R bookdown package, in general.
  
# Specify the location of the bibliography below
bibliography: bib/thesis.bib
# Download your specific csl file and refer to it in the line below.
csl: csl/apa.csl
lot: false
lof: true
---

<!--
Above is the YAML (YAML Ain't Markup Language) header that includes a lot of 
metadata used to produce the document.  Be careful with spacing in this header!

If you'd prefer to not include a Dedication, for example, simply delete the section entirely, or silence them (add # before each line). 

If you have other LaTeX packages you would like to include, delete the # before header-includes and list the packages after hyphens on new lines.

If you'd like to include a comment that won't be produced in your resulting file enclose it in a block like this.

If you receive a duplicate label error after knitting, make sure to delete the index.Rmd file and then knit again.
-->

```{r include_packages, include=FALSE}
# This chunk ensures that the thesisdown package is
# installed and loaded. This thesisdown package includes
# the template files for the thesis.
if (!require(remotes)) {
  if (params$`Install needed packages for {thesisdown}`) {
    install.packages("remotes", repos = "https://cran.rstudio.com")
  } else {
    stop(
      paste('You need to run install.packages("remotes")",
            "first in the Console.')
    )
  }
}
if (!require(thesisdown)) {
  if (params$`Install needed packages for {thesisdown}`) {
    remotes::install_github("ismayc/thesisdown")
  } else {
    stop(
      paste(
        "You need to run",
        'remotes::install_github("ismayc/thesisdown")',
        "first in the Console."
      )
    )
  }
}
library(thesisdown)
# Set how wide the R output will go

```


```{r wrap-hook, echo = FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

```



<!--
The ackowledgements, preface, dedication, and abstract are added into the PDF
version automatically by inputing them in the YAML at the top of this file.
Alternatively, you can put that content in files like 00--prelim.Rmd and
00-abstract.Rmd like done below.
-->



```{r eval=!knitr::is_latex_output(), child=here::here("prelims", "00--prelim.Rmd")}

```

```{r eval=!knitr::is_latex_output(), child=here::here("prelims", "00-abstract.Rmd")}

```

<!-- The {.unnumbered} option here means that the introduction will be 
"Chapter 0." You can also use {-} for no numbers on chapters.
-->

# Introduction {.unnumbered}

If it weren't for Professor Ben Baumer, I would not be here writing this thesis today. The Introduction to Data Science course I took with him in my first semester at Smith College spurred on my love for the field, prompting me to declare a major---with Ben as my advisor---soon after. 

That course---SDS 192---was my first introduction to the topic of scientific reproducibility. In Ben's class, one of the first undergraduate courses in the country to implement reproducibility as part of the curriculum, I was introduced to a variety of tools that could help make my work more reproducible. I learned how to write effective narrative reports with RMarkdown, how to implement version control on GitHub, how to write well-styled code, and how to use R projects to structure my files in a way that was easy to manage. 

Through this work, I began to see the benefits of emphasizing a reproducible workflow. When I worked on projects with other students, the tools I learned in class enabled us to collaborate with one another easily and keep track of the progress we had made. And when I tried to look back at my own work from the past, my attention to reproducibility meant that it was much easier for me to re-compute and understand what I had done previously.

Additionally, my work in Introduction to Data Science helped build a strong interest in R. I developed a love for programming in the language and was curious to learn more about how it worked---both behind the scenes and practice. This interest prompted me to pursue research in the field. I wanted to expand my knowledge and challenge myself by learning about and solving cutting-edge issues in the data science domain. 


As soon as I felt that I had built up enough coding experience to be able to tackle challenging issues, I reached out to the data science faculty to see if there were any opportunities. Ben enthusiastically welcomed me in, telling me that he had recently started developing an R package focused on reproducibility and offering me a position assisting him with writing it. That was when I first learned about `fertile`. 

As part of his interest in reproducibility, Ben had been working---with help from Hadley Wickham and Jenny Bryan of RStudio---to develop a package designed to help users create optimal conditions for reproducible work in R (titled "fertile").  

When it was presented to me, the package was incomplete. Several of the major functions that still exist now were there at the time, but there were not many features beyond the basics. Additionally, several functions were not fully written or throwing errors, the package had little to no documentation, and the majority of tests that had been written to test the functions were failing.

I was given the opportunity to join in on the project and help move `fertile` to a functional state: to fix the errors that were present, document the package, and write new functions to expand the operation beyond the barebones structure that was present.
  
The `fertile` project was an excellent fit for me. The package was not yet in a functional state, but had an excellent base of code to work off of. Given that it already had some basic functions and a strong structure, `fertile` provided a perfect opportunity for me to learn about the process of writing R packages. 

The more I worked on `fertile`, the more interested in the project I became. The package was unlike anything else I had been exposed to in data science. Most of my course work had been in the form of front-end coding: conducting data analyses and visualizing information. `fertile` is nothing like that. It operates completely behind the scenes, conducting a kind of meta-analysis to determine the reproducibility of a data science project. Rather than writing code to analyze data, I had to learn how to write code to analyze code that *other people* wrote to analyze data. 

This was an incredibly complex problem. It involved capturing information about the user's file system and their RStudio environment, both of which I knew nothing about. It also involved tracking and recording user behavior---something definitely not taught in data analysis courses. Even experienced R users deemed some of the goals associated with the package to be incredibly lofty. [On the question](https://stackoverflow.com/questions/43638773/comprehensive-way-to-check-for-functions-that-use-the-random-number-generator-in) of whether it was possible to test whether R code contained uncontrolled randomness, one coder said it was "probably impossible", while another said there was "fundamentally no way" to achieve it.

The challenge of the problems at hand, however, was what made them interesting. Most of the time, there was no way to use my past coding knowledge to solve them---rather, I had to research and test out new solutions. While the process was at times incredibly frustrating, particularly given that some of the errors I came across had likely never been seen by any other R user, it was an incredible learning opportunity. I was exposed to a wide variety of R features I would likely have never seen otherwise and greatly broadened my problem-solving and troubleshooting skills. 

Every problem I solved made me more confident in my abilities and more excited to work on the `fertile`. I delved more into the topic of reproducibility and looked at some of the work by other researchers and code developers. This research introduced me to new aspects of reproducibility I wanted to include in the package and inspired me to expand its scope. 

This paper is the culmination of the work I have been doing with `fertile` for the past two years. 


